---
title: "RERconverge Enrichment Functions Walkthrough"
author: "Amanda Kowalczyk"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
        css: custom.css
        toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This walkthrough provides instructions for implementing the RERconverge package enrichment functions to identify gene pathways or groups of genes whose evolutionary rates shift in association with change in a trait. For information on how to download and install RERconverge, see the [wiki](https://github.com/nclark-lab/RERconverge/wiki/install). Source code and a quick start guide are available on [github](https://github.com/nclark-lab/RERconverge).  A detailed walkthrough to produce the data used for this example is provided in the [vignettes folder](https://github.com/nclark-lab/RERconverge/blob/master/vignettes/RERconvergeAnalysisFullWalkthrough.Rmd).

# Overview

This document describes how to run pathway enrichment analysis using RERconverge output and functions included with the RERconverge package.  Enrichment analysis detects groups of genes that are evolving faster or slower with a phenotype of interest.  In the RERconverge package, the enrichment function is implemented as a Wilcoxon Rank-Sum Test on a list of genes ranked based on their correlation values.  It detects distribution shifts in groups of genes compared to all genes, and it thereby bypasses the need to set a foreground significance cutoff like some other enrichment methods. 
 
**Input** to the enrichment function is the output from RERconverge correlation functions and pathways of interest with gene symbols (gene names). 
 
**Output** is enrichment statistics for each pathway, including genes in the pathways and their ranks.

# Extract Results from RERconverge Correlation Analysis

Enrichment analysis starts with the results from the *correlateWithContinuousPhenotype*, *correlateWithBinaryPhenotype*, or *getAllCor* functions in the RERconverge package.  These results include Rho, p-value, and the Benjamini-Hochberg corrected p-value for the correlation between the relative evolutionary rates of each gene and the phenotype provided.  These statistics are used to calculate enrichments. 
 
In this case, we will start with the data from the continuous trait analysis described in the [full walkthrough](https://github.com/nclark-lab/RERconverge/blob/master/vignettes/RERconvergeAnalysisFullWalkthrough.Rmd) that used adult mass as the phenotype of interest in mammalian species.  This walkthrough assumes that you have already installed RERconverge.

```{r results='hide', message = FALSE, warning = FALSE, cache=T}
library(RERconverge)
data("RERresults")
```

# Convert Statistics

We will perform our enrichment on Rho-signed negative log p-values from our correlation results.  The *getStat* function converts our correlation results to these values and removes NA values.

```{r cache=T}
library(RERconverge)
stats=getStat(res)
```

# Download Pathway Annotations

Now that we have our gene statistics, we need pathway annotations.  Download all curated gene sets, gene symbols (c2.all.v6.2.symbols.gmt) from [GSEA-MSigDB](http://software.broadinstitute.org/gsea/downloads.jsp) as gmtfile.gmt.  You must register for an account prior to downloading.  The rest of the vignette expects "gmtfile.gmt" to be in the current working directory.

# Format Pathway Annotations

RERconverge enrichment functions expect pathways to be in pathway-group lists contained within a list.  Each pathway-group list contains






