---
title: "RERconverge Enrichment Functions Walkthrough"
author: "Amanda Kowalczyk"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This walkthrough provides instructions for implementing the RERconverge package enrichment functions to identify gene pathways or groups of genes whose evolutionary rates shift in association with change in a trait. For information on how to download and install RERconverge, see the [wiki](https://github.com/nclark-lab/RERconverge/wiki/install). Source code and a quick start guide are available on [github](https://github.com/nclark-lab/RERconverge).

# Overview

Download all curated gene sets, gene symbols (c2.all.v6.2.symbols.gmt) from [GSEA-MSigDB](http://software.broadinstitute.org/gsea/downloads.jsp) as gmtfile.gmt.  You must register for 

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
